---
title: "Activity 14: Data Wrangling, Visualization, and Reflection"
author: "Amogh Krishna Namuduri"
date: "11/16/2027"
format: pdf
execute: true
---

# Armed Forces Data Wrangling Redux

## Load packages

Chunk 1

```{r}
# Load required packages 
library(tidyverse) 
library(janitor) 
library(readxl) 

```

## Load the data

Chunk 2

```{r}
# Load the Armed Forces data set
# This file can be accessed on "https://docs.google.com/spreadsheets/d/1cn4i0-ymB1ZytWXCwsJiq6fZ9PhGLUvbMBHlzqG4bwo/edit?usp=sharing"
armed_raw <- read_excel("US_Armed_Forces_(6_2025).xlsx", skip = 1)

```

## View armed_raw

Chunk 3

```{r}
# View head of armed_raw data
head(armed_raw)

```

## Extract header row and view it

Chunk 4

```{r}
header_row <- read_excel("US_Armed_Forces_(6_2025).xlsx", n_max = 1)
header_row

```

## Clean names

Chunk 5

I used clean_names() function from the janitor package. This makes the spaces into underscores. But in some cases renaming was required.

```{r}
cleaned <- armed_raw %>%
  clean_names()
names(cleaned)

```

## Rename column names

Chunk 6

```{r}
cleaned <- armed_raw %>%
  clean_names() %>%
  rename(
    pay_grade = x1,
    army_male = army,
    army_female = x3,
    army_total = x4,
    navy_male = navy,
    navy_female = x6,
    navy_total = x7,
    marine_male = marine_corps,
    marine_female = x9,
    marine_total = x10,
    airforce_male = air_force,
    airforce_female = x12,
    airforce_total = x13,
    space_male = space_force,
    space_female = x15,
    space_total = x16,
    total_overall = total
)

```

## Data wrangling

Chunk 7

I used the pivot_longer() function to reshape the data converting the dataset from a wide format to a long format. In this chunk I also filtered only numeric values leaving the NA columns.

```{r}
group_df <- cleaned |>
  select(pay_grade, matches("_(male|female|total)$")) |>
  pivot_longer(
    cols = -pay_grade,
    names_to = c("branch", "gender"),
    names_sep = "_",
    values_to = "count"
  ) |>
  filter(gender != "total") |>
  mutate(count = as.numeric(count)) |>
  filter(!is.na(count))

```

## Final data set

Chunk 8

```{r}
# Create a clean frequency-style table from group_df
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "The table shows the reshaped data in the columns of Pay Grade, Branch, Gender and Count"
present_df <- group_df |>
  rename(
    Pay_Grade = pay_grade,
    Branch = branch,
    Gender = gender,
    Count = count
  )

kable(present_df, caption = "Reshaped Data: Pay Grade × Branch × Gender")

```

# Baby Names Visualization

For the names I chose Michael and Emily, they are among the most consistently popular names in the US. I can get a trend using these 2 names which would allow me to explore the long-term popularity and compare how the trends have changed for male and female names. The plot I choose is Line plot, this is because it will give us a comparison of the trend between the 2 names from 1880 to the present.

Chunk 9

```{r}
#| fig-pos: "H"
#| fig-width: 10
#| fig-cap: "The plot shows us the Popularity of the names Michael and Emily over time"
library(babynames)
library(dplyr)
library(ggplot2)
selected_names <- babynames %>%
  filter(name %in% c("Michael", "Emily"))
ggplot(selected_names, aes(x = year, y = n, color = name, linetype = name, group = interaction(name,sex))) +
  geom_line(size = 1) +
  labs(
    title = "Popularity of Michael and Emily Over Time",
    x = "Year",
    y = "Number of Babies",
    color = "Name",
    linetype = "Name"
    ) +
    theme_minimal()

```

# Plotting a Mathematical Function

For calculating the volume of a box with respect to the side length that is cut out, we will have to use the below function

V(x) = x \* (30-2x) \* (20-2x)

The code below is for the plot that shows how the Volume is affected when the value of x increases

Chunk 10

```{r}
#| fig-pos: "R"
#| fig-cap: "The plot shows how the Volume changes as the value of x changes"
# Define the volume function
volume <- function(x) {
  x * (30 - 2*x) * (20 - 2*x)
}
# Create a sequence of x values from 0 to 10
x_vals <- seq(0, 10, length.out = 300)
vol_vals <- volume(x_vals)
# Plot
library(ggplot2)
ggplot(data.frame(x = x_vals, volume = vol_vals), aes(x, volume)) +
  geom_line(color = "steelblue", linewidth = 1.3) +
  labs(
    title = "Box Volume as a Function of Cut Size x",
    x = "Cut Size (x)",
    y = "Volume"
  ) +
  theme_minimal()

```

# What I feel I have learned

I have learnt that there are many things to take into consideration when we want to make our code or work open to all and reproducible, such as it should be able to run on anyone's workstation. Based on learning in R, I have learnt that it is best to keep rechecking your code so that it follows all the steps of data wrangling so that we get the best output. I have learnt how to use clean_name() and uncount() in a proper sense. I have come from a state of understnading the code I am using to writing it part by part by myself.

# Code Appendix

```{r}
# Load required packages 
library(tidyverse) 
library(janitor) 
library(readxl) 

```

```{r}
# Load the Armed Forces dataset
armed_raw <- read_excel("US_Armed_Forces_(6_2025).xlsx", skip = 1)

```

```{r}
# View head of armed_raw data
head(armed_raw)

```

```{r}
header_row <- read_excel("US_Armed_Forces_(6_2025).xlsx", n_max = 1)
header_row

```

```{r}
cleaned <- armed_raw %>%
  clean_names()
names(cleaned)

```

```{r}
cleaned <- armed_raw %>%
  clean_names() %>%
  rename(
    pay_grade = x1,
    army_male = army,
    army_female = x3,
    army_total = x4,
    navy_male = navy,
    navy_female = x6,
    navy_total = x7,
    marine_male = marine_corps,
    marine_female = x9,
    marine_total = x10,
    airforce_male = air_force,
    airforce_female = x12,
    airforce_total = x13,
    space_male = space_force,
    space_female = x15,
    space_total = x16,
    total_overall = total
)

```

```{r}
group_df <- cleaned |>
  select(pay_grade, matches("_(male|female|total)$")) |>
  pivot_longer(
    cols = -pay_grade,
    names_to = c("branch", "gender"),
    names_sep = "_",
    values_to = "count"
  ) |>
  filter(gender != "total") |>
  mutate(count = as.numeric(count)) |>
  filter(!is.na(count))

```

```{r}
# Create a clean frequency-style table from group_df
#| fig-pos: "H"
#| fig-width: 5
#| fig-alt: "The table shows the reshaped data in the columns of Pay Grade, Branch, Gender and Count"

present_df <- group_df |>
  rename(
    Pay_Grade = pay_grade,
    Branch = branch,
    Gender = gender,
    Count = count
  )

kable(present_df, caption = "Reshaped Data: Pay Grade × Branch × Gender")

```

```{r}
#| fig-pos: "H"
#| fig-width: 10
#| fig-cap: "The plot shows us the Popularity of the names Michael and Emily over time"

library(babynames)
library(dplyr)
library(ggplot2)
selected_names <- babynames %>%
  filter(name %in% c("Michael", "Emily"))
ggplot(selected_names, aes(x = year, y = n, color = name)) +
  geom_line(size = 1) +
  labs(
    title = "Popularity of Michael and Emily Over Time",
    x = "Year",
    y = "Number of Babies",
    color = "Name"
    ) +
    theme_minimal()

```

```{r}
#| fig-pos: "R"
#| fig-cap: "The plot shows how the Volume changes as the value of x changes"

# Define the volume function
volume <- function(x) {
  x * (30 - 2*x) * (20 - 2*x)
}
# Create a sequence of x values from 0 to 10
x_vals <- seq(0, 10, length.out = 300)
vol_vals <- volume(x_vals)
# Plot
library(ggplot2)
ggplot(data.frame(x = x_vals, volume = vol_vals), aes(x, volume)) +
  geom_line(color = "steelblue", linewidth = 1.3) +
  labs(
    title = "Box Volume as a Function of Cut Size x",
    x = "Cut Size (x)",
    y = "Volume"
  ) +
  theme_minimal()

```

